---
title: 'BISC 307, Lab 2: NEON Megapit Soil Profiles'
author: "Prof. Jackie Hatala Matthes, FA 2020"
date: 'Lab: 5-6 November 2020'
output:
  html_document: default
  pdf_document: default
---
[NEON Megapit sampling design](https://data.neonscience.org/data-products/DP1.00096.001)

```{r, results='hide', message=FALSE, warning=FALSE}
#install.packages("neonUtilities")

# Set global option to NOT convert all character variables to factors, which helps to make the NEON functions to work best
options(stringsAsFactors=F)

# Load required packages
library(neonUtilities)
library(tidyverse)

#Read in all available Megapit data
Megapit_data <- loadByProduct(dpID = "DP1.00096.001", check.size = F)

# Unlist Megapit_data to Rstudio environment 
list2env(Megapit_data, .GlobalEnv)
```

* the horizon name and depth data in`mgp_perhorizon`
* biogeochemistry data in `mpg_biogeosample`
* bulk density data in `mgp_perbulksample`

```{r}
# Depths of horizon layers and horizon IDs per site
# Use mutate to create a horizon name that is just the first letter
soil_horizons <- mgp_perhorizon %>%
  select(-uid, -remarks, -publicationDate) %>% # minus sign "unselects"
  mutate(horizonNameSimple = str_extract(horizonName, "[A-Z]{1}"))

soil_horizons = subset(soil_horizons, select = c(horizonID, siteID, pitID, setDate, collectDate, domainID, horizonName, horizonTopDepth, horizonBottomDepth))

# Soil texture, elemental, and nutrient data per horizon
soil_biogeochem <- mgp_perbiogeosample %>%
  select(-uid, -remarks) %>%
  filter(biogeoSampleType=="Regular")

soil_biogeochem = subset(soil_biogeochem, select = c(siteID, horizonID, pitID, horizonName, setDate, collectDate, domainID, carbonTot))
```

```{r}
# Join biogeochemical and physical data 
# function structure: left_join(df_left, df_right, by = c(set of columns in common) - i.e., the key)
soils <- left_join(soil_horizons, soil_biogeochem, 
               by=c('horizonID', 'siteID', 'pitID','setDate',
                    'collectDate', 'domainID', 'horizonName')) %>%
  arrange(siteID, horizonBottomDepth) # Sort by siteID then horizon depth
```


```{r}
# Load csv file with NEON site data
neon_sites <- read_csv("NEON-field-sites.csv")

# Attach NEON site data to soil profiles
soils_sites <- soils %>%
  left_join(neon_sites, by = c("domainID","siteID"))
```


This could be modified to only show what I need for the final project
```{r}
rootsites<-filter(soils_sites, 
siteID =="HARV" | siteID == "UNDE" | siteID == "YELL")

```
***

This code framework may come in handy
```{r}
# Data processing steps:  
# 1. Filter to just the O-horizon data
# 2. Use mutate to calculate O-horizon depth from top/bottom difference
# 3. Plot O-horizon depth vs mean annual precipitation
# 4. Linear regression: O-depth ~ precipitation (y ~ x)

# Steps 1 & 2: Data Wrangling
Ohorizon_precip <- soils_sites %>%
  filter(horizonNameSimple == "O") %>%
  mutate(horizonDepth = horizonBottomDepth - horizonTopDepth)
  
# Step 3: Plot
ggplot(Ohorizon_precip, aes(x = meanAnnualPrecip_mm, y = horizonDepth)) +
  geom_point() +
  labs(x = "Mean Annual Precipitation (mm)", y = "O-horizon depth (cm)") +
  theme_bw()
```
Framework for log plotting below
```{r}
# Step 3: Plot with log(horizonDepth)
ggplot(Ohorizon_precip, aes(x = meanAnnualPrecip_mm, y = log(horizonDepth))) +
  geom_point()+
  labs(x = "Mean Annual Precipitation (mm)", y = "ln O-horizon depth (cm)") +
  theme_bw()

```

double-log plot below:
```{r}
# Step 3: Plot with log(horizonDepth) ~ log(precip)
ggplot(Ohorizon_precip, aes(x = log(meanAnnualPrecip_mm), y = log(horizonDepth))) +
  geom_point()+
  geom_smooth(method = "lm") +
  labs(x = "ln Mean Annual Precipitation (mm)", y = "ln O-horizon depth (cm)") +
  theme_bw()

```


```{r}
# 4a. Calculate correlation between log(horizonDepth) & log(meanAnnualPrecip_mm)
cor.test(log(Ohorizon_precip$horizonDepth), 
         log(Ohorizon_precip$meanAnnualPrecip_mm))

```
From this output, we can see that the correlation coefficient (called `r` when you're writing it in a paper) is equal to -0.75, indicating a strong negative correlation between these two variables. Closer to the top of the output we also see that the p-value < 0.01, which indicates this correlation is signfiicant at the alpha = 0.01 value.

Assessing whether the slope of a linear regression model is different from zero is a stronger test of statistical inference. This test assumes that the x-axis is a predictor variable (indpendent variable) for the data on the y-axis (dependent variable). We can estimate a linear regression model for data in R using the `lm()` function from base R:

```{r}
# 4b. Create linear regression model for log(O-depth) ~ log(Precip)  
# *** The format for the lm() function is y-variable ~ x-variable
reg_OdepthMAP <- lm(log(horizonDepth) ~ log(meanAnnualPrecip_mm), 
                    data = Ohorizon_precip)

# Look at linear regression summary
summary(reg_OdepthMAP)
     ```


